---
title: "Multinomial Logit Model: A Discrete Choice Analysis"
author: "Pushya Shree Konasale Jayaramu"
output: html_document
---

# Introduction  

This project implements a **Multinomial Logit Model (MNL)** to analyze discrete choices among transportation modes. The model estimates mode choice probabilities based on individual and alternative-specific attributes.  

---

## Required Libraries  

```{r, message=FALSE, warning=FALSE}
# Install necessary packages if not installed
if (!require("mlogit")) install.packages("mlogit", dependencies=TRUE)
if (!require("corrplot")) install.packages("corrplot", dependencies=TRUE)
if (!require("ggplot2")) install.packages("ggplot2", dependencies=TRUE)
if (!require("reshape2")) install.packages("reshape2", dependencies=TRUE)

# Load libraries
library(mlogit)
library(corrplot)
library(ggplot2)
library(reshape2)
```

## Data

The dataset contains individual characteristics (age, gender, income) and mode-specific attribute (travel time). Two datasets are used for each region:

1. Hersching
- **Training Data:** `1.h.csv`  
- **Test Data:** `1.h.test.csv`

2. Moosach
- **Training Data:** `1.m.csv`  
- **Test Data:** `1.m.test.csv`

## Read and Inspect Data
```{r}
# Read data
setwd("F:/onedrive/Mode Choice")
fit_data <- read.csv("1.h.csv", sep = ",")

# Summary statistics
summary(fit_data)

# Check unique mode choices
unique(fit_data$mode)
prop.table(table(fit_data$mode))

```

## Correlation Analysis
```{r, fig.width=10, fig.height=6}
# Check correlation among key variables
cormtx <- cor(data.frame(fit_data$age, fit_data$gender, fit_data$income, fit_data$auto_avail))
corrplot(cormtx, method = "number")

```

## Visualizing Income Distribution Across Modes
```{r, fig.width=10, fig.height=6}
ggplot(fit_data, aes(mode, income)) + 
  geom_boxplot() + 
  xlab("Mode") + 
  ylab("Income Distribution")

```


## Convert Data to mlogit Format
```{r}
logit_data <- mlogit.data(fit_data, choice = "mode", shape = "wide", varying = 8:11)
View(as.data.frame(logit_data))

```

## Estimate Multinomial Logit Models

```{r, results="hide"}
# Model 1: Base model with travel time
mnl_fit1 <- mlogit(mode ~ 0 | auto_avail | ttime, data = logit_data, weights = weight)
summary(mnl_fit1)

# Model 2: Mode choice based on age
mnl_fit2 <- mlogit(mode ~ ttime | age | 0, data = logit_data, weights = weight, reflevel = "evtol")
summary(mnl_fit2)

# Model 3: Age and income as person-specific variables
mnl_fit3 <- mlogit(mode ~ 0 | age + income | ttime, data = logit_data, weights = weight)
summary(mnl_fit3)

# Model 4: Adding auto availability
mnl_fit4 <- mlogit(mode ~ 0 | age + income + auto_avail | ttime, data = logit_data, weights = weight)
summary(mnl_fit4)

# Try and fit number of models based on field specific knowledge
```


## Add Binary Variables for Modes
```{r, results="hide"}
logit_data$transit <- ifelse(logit_data$alt=="transit", 1, 0)
logit_data$taxi <- ifelse(logit_data$alt=="taxi", 1, 0)
logit_data$evtol <- ifelse(logit_data$alt=="evtol", 1, 0)
logit_data$car <- ifelse(logit_data$alt=="car", 1, 0)

```

## Final Multinomial Logit Models
```{r}
# Final Model for Moosach: Including interaction terms
#mnl_fit12 <- mlogit(mode ~ I(age*taxi) + I(age*evtol) + I(income*evtol) + 
#                     I(ttime*evtol) + I(ttime*taxi) + I(ttime*transit) | 
#                     auto_avail | 0, data = logit_data, weights = fit_data$weight)
#summary(mnl_fit12)

# Final Model for Hersching: Different interactions
mnl_fit13 <- mlogit(mode ~ I(income*transit) + I(age*taxi) + I(age*evtol) + 
                     I(auto_avail*transit) + I(ttime*evtol) + I(ttime*taxi) | 
                     1 | 0, data = logit_data, weights = weight)
summary(mnl_fit13)

```

## Extracting Model Coefficients
```{r}
mnl_coeff <- data.frame(coefficients(mnl_fit13))
View(mnl_coeff)

```

## Comparing Predicted vs Actual Mode Shares
```{r}
# Predicted mode shares
model <- apply(predict(mnl_fit13, newdata = logit_data), 2, weighted.mean, w = fit_data$weight)

# Actual mode shares
actual <- prop.table(xtabs(weight ~ mode, data = fit_data))

# Combine results
combine_fits <- cbind(model, actual)
combine_fits

```

## Predict Mode Choice on Test Data
```{r}
# Read test dataset
setwd("F:/onedrive/Mode Choice")
test_data <- read.csv("1.h.test.csv")

# Convert test dataset
logit_test_data <- mlogit.data(test_data, choice = "mode", shape = "wide", varying = 8:11)

# Add binary mode indicators
logit_test_data$evtol <- ifelse(logit_test_data$alt=="evtol", 1, 0)
logit_test_data$transit <- ifelse(logit_test_data$alt=="transit", 1, 0)
logit_test_data$car <- ifelse(logit_test_data$alt=="car", 1, 0)
logit_test_data$taxi <- ifelse(logit_test_data$alt=="taxi", 1, 0)

# Predict mode shares on test data
predict(mnl_fit13, newdata = logit_test_data, weights = test_data$weight)

```

## Compare Predictions
```{r}
# Aggregated predictions
test_predicted <- apply(predict(mnl_fit13, newdata = logit_test_data), 2, weighted.mean, w = test_data$weight)
test_actual <- prop.table(xtabs(weight ~ mode, data = test_data))

# Combine results
combine_fits <- cbind(combine_fits, test_predicted, test_actual)
combine_fits

```

## Visualizing Actual vs Predicted Mode Shares
```{r, fig.width=10, fig.height=6}
# Reshape data for plotting
melted_data <- melt(combine_fits)
names(melted_data) <- c("mode", "dataset", "value")

# Plot comparison
ggplot(melted_data, aes(mode, value)) + 
  geom_bar(aes(fill = dataset), position = "dodge", stat = "identity") + 
  xlab("Mode") + 
  ylab("Percent Share") +
  ggtitle("Predicted vs Actual Mode Shares")

```

## Conclusion
1. The Multinomial Logit Model successfully estimated mode choices based on age, income, travel time, and auto availability.
2. Interaction effects helped reveal differences in mode preference among individuals.
3. Model predictions were validated against a test dataset, with reasonable alignment to actual mode shares.
4. Further improvements could explore Nested Logit Models or additional attributes like cost and convenience.
